---
title: "Design decisions worth arguing with"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design decisions worth arguing with}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(edc2cdisc)
```

This vignette is the package's opinion column: the decisions behind the
mappers, why they are the way they are, and where they are open to
argument. The package itself is deliberately opinionated — a mapping is a
series of decisions, and hiding them behind cleverness is how SDTM projects
rot.

## Everything is character on read

`read_clinical_view()` reads every column as character. EDC exports quote
all values, mix numeric and text across CRF versions, and carry partial
dates no parser should be allowed to guess at. Typing happens later,
explicitly, per variable (`numeric`, `dtc`, `yn` transforms in the spec).
The cost is a stream of `as.numeric()` calls in the mappers; the benefit is
that nothing is ever typed by accident.

## Partial dates stay partial in SDTM

Rave splits dates into year/month/day components; blanks mean UNK. That is
exactly ISO 8601 reduced precision, so `rave_dtc()` builds `2024-03` or
`2024` and stops. SDTM keeps the precision it was given.

Imputation is an analysis-layer decision, and it happens exactly once:
`impute_dtc()` fills missing month/day with 01 (the "first" rule, also
admiral's default) and returns a flag (`"D"`, `"M"`) recording what was
filled. ADSL, ADAE, ADVS and ADLB all consume that one rule.

The bias is real: imputing a birth date to January 1st can overstate AGE by
up to 11 months. A study SAP often prefers mid-year. The point of
`impute_dtc()` is not that the first rule is right — it is that there is
one explicit, flagged rule to argue with, rather than silence.

For the same reason, reduced-precision ISO dates compare correctly as
strings (`"2023" < "2024-03-17"`), which is how the MH-before-first-dose
validator works without imputing anything.

## One spec, two readers

The mapping configuration lives in a `study_spec` (see
`vignette("new-study")`). The `variables` table names a small transform
vocabulary — `rename`, `decode`, `dtc`, `yn`, `numeric`, `verbatim`,
`constant` — and a derivation registry for everything semantic. The engine
that applies it is tiny; `map_variables()` is basically `transmute()`
driven by the table.

What makes this more than documentation is that **the validators read the
same spec**: the required-variable lists for the engine-mapped domains come
straight from `spec$variables`, and the decodes go through
`spec$codelists`. A spec that silently lies breaks the build. Two things
read it.

## Codelists fail loudly

Every controlled-terminology mapping goes through `check_ct()`: an
unmapped collected value **stops the run** rather than passing through as
NA. When the medical monitor adds a new severity term, you get an error
naming the variable and the value, not a silently truncated AESEV. The
flip side: mapping a new value is always a deliberate edit to the spec.

## Screen failures have no study days

A screen failure never has an RFSTDTC, so every `--DY` derivation is NA by
construction — and the validator enforces that nobody leaked a study day
onto a screen-failure subject. DM ages them at informed consent instead of
first dose; ADaM carries their arm as absent rather than a placeholder.
The VISITDY exception is explicit: it is a protocol-planned day, not
anchored on treatment.

## Deaths must agree three ways

A fatal AE outcome, the DS DEATH disposition record, and the collected DM
death fields are three independent claims about the same fact. The
validators require all three to name the same subjects, the dates to
agree, and the death to anchor ADSL's LSTALVDT and discontinuation reason.
Deaths are also the case a strict treatment-emergent window keeps: onset
before the last dose, outcome after it.

## TRTEMFL is an ADaM variable

SDTM keeps the collected facts; the treatment-emergent flag is applied in
`derive_adae()`, recomputed from ADSL dates by `validate_adam()`. The
window is onset within `[TRTSDT, TRTEDT]` — no grace period, because no
SAP in this project defines one. If your study defines a grace window,
that is one `case_when()` to edit, and the validator will follow you.

## SUPP drops blanks; CO is for prose

`make_supp()` emits a SUPP row only where the qualifier value is present:
a "specify" field appears only for subjects who triggered it. Free-text
investigator comments go to CO, not SUPP — structured name/value pairs
versus narrative tied to a record.

## Trial design is pure spec

TA, TE, TI, TV and TS never touch the mapping engine. Four optional spec
tables hold the protocol; TV is derived from the `visits` table the
scheduled-visit domains already read. The validators re-read the same
tables — NARMS against `spec$arms`, PLANSUB against `spec$study$n`, TV
against `spec$visits`, SV against TV — so a spec that lies about its own
design cannot pass both readers, which is the same discipline the ADaM
checks apply to the SDTM layer.

## The honest limit

"A new study is a spec change" is true for studies in the same CRF family:
same forms, different codelists, arms, visits, sites, SUPP fields. It is
**not** true when the CRF design changes. "Which field means dose
administered" or "screen failure arrives on the protocol-deviation code"
are semantics, and semantics live in the derivation registry as code that
the spec *selects*. That boundary is a feature: new study = new spec plus,
rarely, one or two new derivation functions. Over-promising fully
metadata-driven SDTM is how these projects collapse.
