---
title: "A new study is a spec change"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A new study is a spec change}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(edc2cdisc)
```

Everything that changes between studies — sites, arms, visits, codelists,
SUPP qualifiers, unit conversions — lives in one `study_spec` object. The
shipped `spec_synth01` is the reference implementation for the synthetic
study the package generates; `spec_synth02` is a second, complete worked
example — different sites, arms, visit schedule, codelist decodes, an extra
vital sign and an extra SUPP qualifier — that runs through `build_all()`
with no mapper changes. This vignette walks through the kinds of
changes a new study in the same CRF family needs, and where the boundary
of "just a spec change" sits.

## What a spec holds

```{r}
spec_synth01
names(spec_synth01)
```

Each table carries one kind of configuration: `study` (constants),
`sites`, `arms`, `visits`, `codelists` (collected decode -> CDISC term),
`supp` (SUPP-- qualifiers), `units` (LB conventional-to-SI factors),
`forms` (event vs log typing), `tests` (per-test pivot specs for VS/LB)
and `variables` (one row per SDTM variable per domain: source CRF field,
transform, derivation reference).

## Changing a controlled-terminology mapping

Suppose the sponsor redefines causality so that only "Related" counts as
related. Edit the codelist rows and everything downstream follows — the
AE mapper decodes through `ct_lookup()`, so an unmapped value fails loudly
rather than drifting:

```{r, eval = FALSE}
spec2 <- spec_synth01
spec2$codelists$cdisc_term[spec2$codelists$ct == "AEREL"] <-
  c("NOT RELATED", "NOT RELATED", "NOT RELATED", "RELATED")
```

## Adding a site, an arm, a visit

Sites, arms and visits are plain rows. A study adds its own:

```{r, eval = FALSE}
spec3 <- spec_synth01
spec3$sites <- rbind(
  spec3$sites,
  tibble::tribble(
    ~siteid, ~SiteNumber, ~Site, ~SiteGroup, ~StudySiteId, ~COUNTRY,
    "4104",  "104",       "New Site General", "Europe", "7704", "FRA"
  )
)
```

The DM mapper's country derivation reads `spec$sites`, so COUNTRY follows
without touching code — and `validate_sdtm()` derives its required lists
from the same object.

## Adding a SUPP qualifier

A new non-standard CRF field is one row in `spec$supp`: which domain, the
IDVAR pinning, the QNAM/QLABEL, the raw source column and the transform
that produces QVAL:

```{r, eval = FALSE}
spec4 <- spec_synth01
spec4$supp <- rbind(
  spec4$supp,
  tibble::tribble(
    ~rdomain, ~idvar,   ~qnam,   ~qlabel,        ~src,    ~transform, ~qorig, ~qeval,
    "EX",     "EXSEQ",  "EXLOT", "Lot Number",   "EXLOT", "squish",   "CRF",  NA
  )
)
```

`map_suppex()` picks the row up; blank values still produce no SUPP
record, exactly as in the shipped study.

## Adding a vital sign

The findings domains pivot off `spec$tests` — one row per test, not per
variable. Adding a respiratory rate collection is a row, not a code
change:

```{r, eval = FALSE}
spec5 <- spec_synth01
spec5$tests <- rbind(
  spec5$tests,
  tibble::tribble(
    ~domain, ~field,  ~testcd, ~test,                ~cat, ~specimen,
    "VS",    "RESP",  "RESP",  "Respiration Rate",   NA,   NA
  )
)
```

If the parameter should reach ADVS, declare its analysis configuration in
`spec$bds` — order and reference ranges. A parameter built into ADVS
without a `spec$bds` row is caught by the validator's coverage check, so
the CRF addition cannot silently skip the analysis layer:

```{r, eval = FALSE}
spec5$bds <- rbind(
  spec5$bds,
  tibble::tribble(
    ~domain, ~paramcd, ~paramn, ~anrlo, ~anrhi,
    "ADVS",  "RESP",   7,       12,     20
  )
)
```

## Building with your spec

Any spec variant runs through the same pipeline:

```{r, eval = FALSE}
built <- build_all("path/to/extract", spec = spec5)
```

The validators read your spec too, so required-variable checks stay in
sync with what the mappers emit. For a complete, tested example of a
second study built exactly this way, see `spec_synth02`: generating its
extract and running `build_all(extract, spec = spec_synth02)` passes both
validators with zero mapper changes, and feeding that extract to
`build_all(extract, spec = spec_synth01)` fails loudly — the spec really
is what drives the pipeline.

## The boundary

What a spec change **cannot** do is change semantics: which field means
"dose administered", how a screen failure is recognised, what counts as
treatment-emergent. Those live in the derivation registry — named
functions the spec selects — and a study that changes the CRF design adds
or edits one or two of them. That is the honest limit; see
`vignette("design")` for why over-promising fully metadata-driven SDTM is
how these projects collapse.
